Create sequence of ggplots and send to pdf
========================================================

Analysis background
--------------------

See my previously posted tutorial [Getting started with R, with plyr and ggplot2, for group by analysis](http://hselab.org/getting-started-R-group-by.html). At the end of that tutorial we did some facet plots to show a bunch of histograms. What if we wanted to do a sequence of plots instead and save each of them to separate PDFs?

One approach
Let's get ggplot2 loaded…

```{r}
library(ggplot2)
```
and the data file read in.

```{r}
sched_df <- read.csv("data/SchedDaysAdv.csv")
```


Here's the original facet plot of histograms of schedule lead time by service.

```{r}
# Histogram with frequencies
ggplot(sched_df, aes(x = ScheduledDaysInAdvance)) + facet_wrap(~Service) + geom_histogram(aes(y = ..density..), 
    binwidth = 4)
```

We want a sequence of plots and to save each to its own PDF. Some Googling and StackOverflowing led me to the following approach using a custom function and lapply() for the “looping”.

```{r}
doPlot = function(svc_name) {
    # Just get the records for this service
    temp_df = subset(sched_df, Service == svc_name)

    # Create the plot object
    ggobj = ggplot(temp_df, aes(x = ScheduledDaysInAdvance)) + geom_histogram(aes(y = ..density..), 
        binwidth = 4)

    # Add some titles and axis labels
    ggobj = ggobj + ggtitle(svc_name) + xlab("Scheduled Days in Advance") + 
        ylab("Frequency")

    # Need to use print() to actually produce the plot
    print(ggobj)

    # Save to PDF
    ggsave(sprintf("%s.pdf", svc_name))
}
```


Now we can use lapply() to apply our function to a list of unique service names. We can get those unique names with the unique() function. Done.

```{r}
lapply(unique(sched_df$Service), doPlot)
```


